Black swans in ecological time series\\Black swans in population dynamics\\Heavy-tailed ecological time series\\Black swans in ecological dynamics

### 2014-10-17 supp.tex
SOM tweaks

\begin{enumerate}
\item
  To remove populations with unreliable population indices that could be strongly confounded with economics and sampling effort, we removed all populations with a sampling protocol listed as \texttt{harvest} as well populations with the words \texttt{harvest} or \texttt{fur} in the cited reference title.
\item
  We removed all populations with uneven sampling intervals. I.e.~we removed populations that didn't have a constant difference between the ``decimal year begin'' and ``decimal year end'' columns.
\item
  We removed all populations rated as $< 2$ in the GPDD quality assessment (on a scale of $1$ to $5$, with $1$ being the lowest quality data) \citep[following][]{sibly2005, ziebarth2010}
\item
  Populations with negative abundance values were assumed to be log values and transformed by taking $10$ to the power of the recorded abundance. In many cases this was noted for the population, but not in all cases. We inspected each of these \totalAssumedLog~time series to make sure our assumption made sense.
\item
  We filled in all missing time steps with \texttt{NA} values and imputed single missing values with the geometric mean of the previous and following values. We chose a geometric mean to be linear on the log scale that the Gompertz and Ricker-logistic models were fitted on.
\item
  We filled in single recorded values of zero with the lowest non-zero value in the time series \citep[following][]{brook2006a}. This assumes that single values of zero result from abundance being low enough that censusing missed present individuals. We turned multiple zero values in a row into \texttt{NA} values. This implies that multiple zero values were either censusing errors or caused by emigration. Regardless, our population models were fitted on a multiplicative (log) scale and so could not account for zero abundance.
\item
  We removed all populations with four or more identical values in a row since these suggest either recording error or extrapolation between two observations.
\item
  We removed all populations without at least four unique values \citep[following][]{brook2006a}.
\item
  We then wrote an algorithm to find the longest unbroken window of abundance (no \texttt{NA}s) with at least 20 time steps in each population time series. If there were any populations with multiple windows of identical length, we took the most recent window. This is a longer window than used in some previous analyses \citep[e.g.][]{brook2006a}, but since our model attempts to capture the shape of the distribution tails, our model requires more data.
\item
  Finally, we removed GPDD Main ID \texttt{20531}, which we noticed was a duplicate of \texttt{10139} (a heron population).
\end{enumerate}

We provide a supplemental figure of all the time series included in our analysis and indicate which values were interpolated (\percImputedPops\% of populations had at least one point interpolated and only \percImputedPoints\% of the total observations were interpolated) (Fig.~\ref{fig:all-ts}). Table S1 shows the final taxonomic breakdown and the number of populations with interpolated values.

@@

First, we tested the ability to estimate $\nu$ given different true values of $\nu$ and sample sizes. We took stochastic draws from t distributions with different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$ [$\approx$ normal]), with central tendency parameters of $0$, and scale parameters of $1$. We started with 1600 stochastic draws and then fitted again at the first 800, 400, 200, 100, 50, and 25 draws. Each time we recorded the posterior samples of $\nu$.

@@

We considered a version of the Gompertz model in which an autoregressive parameter was fitted to the process noise residuals: 
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma).
\end{align*}
In addition to the parameters in the original Gompertz model, we estimate an additional parameter $\phi$, which represents the relationship between subsequent residuals. Based on the results of previous analyses with the GPDD \citep[e.g.][]{connors2014} and the chosen priors in previous analyses \citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on $\phi$ that assumed the greatest probability density near zero with the reduced possibility of $\phi$ being near $-1$ or $1$. Specifically, we chose $\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. The MCMC chains for a small number of these models (XX) did not converge according to our criteria ($\widehat{R} < 1.05, n_\mathrm{eff} > 200$) after 8000 iterations of four chains. We do not show these models in Fig.~\ref{fig:alt}.

>>>>

\begin{enumerate}
\item
  To remove populations with unreliable population indices that could be strongly confounded with economics and sampling effort, we removed all populations with a sampling protocol listed as \texttt{harvest} as well populations with the words \texttt{harvest} or \texttt{fur} in the cited reference title.
\item
  We removed all populations with uneven sampling intervals. I.e.~we removed populations that didn't have a constant difference between the ``decimal year begin'' and ``decimal year end'' columns.
\item
  We removed all populations rated as $< 2$ in the GPDD quality assessment (on a scale of $1$ to $5$, with $1$ being the lowest quality data) \citep[following][]{sibly2005, ziebarth2010}
\item
  Populations with negative abundance values were assumed to be log values and transformed by taking $10$ to the power of the recorded abundance. In many cases this was noted for the population, but not in all cases. We inspected each of these \totalAssumedLog~time series to make sure our assumption made sense.
\item
  We filled in all missing time steps with \texttt{NA} values and imputed single missing values with the geometric mean of the previous and following values. We chose a geometric mean to be linear on the log scale that the Gompertz and Ricker-logistic models were fitted on.
\item
  We filled in single recorded values of zero with the lowest non-zero value in the time series \citep[following][]{brook2006a}. This assumes that single values of zero result from abundance being low enough that censusing missed present individuals. We turned multiple zero values in a row into \texttt{NA} values. This implies that multiple zero values were either censusing errors or caused by emigration. Regardless, our population models were fitted on a multiplicative (log) scale and so could not account for zero abundance.
\item
  We removed all populations with four or more identical values in a row since these suggest either recording error or extrapolation between two observations.
\item
  We removed all populations without at least four unique values \citep[following][]{brook2006a}.
\item
  We then wrote an algorithm to find the longest unbroken window of abundance (no \texttt{NA}s) with at least 20 time steps in each population time series. If there were any populations with multiple windows of identical length, we took the most recent window. This is a longer window than used in some previous analyses \citep[e.g.][]{brook2006a}, but since our model attempts to capture the shape of the distribution tails, our model requires more data.
\item
  Finally, we removed GPDD Main ID \texttt{20531}, which we noticed was a duplicate of \texttt{10139} (a heron population).
\end{enumerate}
\noindent
We provide a supplemental figure of all the time series included in our analysis and indicate which values were interpolated (\percImputedPops\% of populations had at least one point interpolated and only \percImputedPoints\% of the total observations were interpolated) (Fig.~\ref{fig:all-ts}). Table S1 shows the final taxonomic breakdown and the number of populations with interpolated values.

@@

First, we tested the ability to estimate $\nu$ given different true values of $\nu$ and sample sizes. We took stochastic draws from t distributions with different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$ [$\approx$ normal]), with central tendency parameters of $0$, and scale parameters of $1$. We started with 1600 stochastic draws and then fitted the models again at the first 800, 400, 200, 100, 50, and 25 draws. Each time we recorded the posterior samples of $\nu$.

@@

We considered a version of the Gompertz model in which an autoregressive parameter was fitted to the process noise residuals: 
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma).
\end{align*}
In addition to the parameters in the original Gompertz model, we estimate an additional parameter $\phi$, which represents the relationship between subsequent residuals. Based on the results of previous analyses with the GPDD \citep[e.g.][]{connors2014} and the chosen priors in previous analyses \citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on $\phi$ that assumed the greatest probability density near zero with the reduced possibility of $\phi$ being near $-1$ or $1$. Specifically, we chose $\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. The MCMC chains for a small number of these models (XX) did not converge according to our criteria ($\widehat{R} < 1.05, n_\mathrm{eff} > 200$) after 8000 iterations of four chains. We did not include these models in Fig.~\ref{fig:alt}.
